# Command line specification

## Rationale

In an effort to provide a maximally compatible `argparse` replacement that does
not rely on Python, this document aims to specify a minimal set of supported
command line argument options. This should avoid feature creep and remain
quickly implementable — and optionally extensible. On the other hand, it should
provide all or most of the standard Unix command line syntax, and offer large
compatibility with it.

The API is intentionally not identical to `argparse`’s (pythonic) API. Instead,
an effort is made to provide an opinionated API familiar to R users.

## Supported command line syntax

* Supported are both short and long named arguments and positional arguments,
  all of which may be declared as optional and mandatory, and assigned default
  values
* Short arguments are preceded by `-` and may be grouped into one long string
* Long arguments are preceded by `--`
* Values of short arguments are separated by whitespace or directly adjacent
* Values of long arguments are separated by whitespace, assignment (`=`) or both
* Positional arguments follow named arguments. The special argument `--`
  (without a following name) specifies that all following arguments are
  positional, even if they start with one or more dashes
* Even if undeclared, the special named argument `--help` (`-h`) will print a
  summary of the available command line arguments to stderr and exit
* The same happens when an invalid argument is recognised, but in this case
  additionally with a nonzero exit status

## API

Command line arguments are specified via `opt`:

```{r}
opt = function (short, long, default, description, validation) {
    # FIXME: Make default optional, which makes `opt` mandatory
    stopifnot(is.character(short) && length(short) == 1)
    stopifnot(is.character(long) && length(long) == 1)
    stopifnot(length(default) == 1)
    stopifnot(is.character(description) && length(description) == 1)

    if (is.null(names(default))) {
        stopifnot(long != '')
        variable_name = long
    }
    else
        variable_name = names(default)

    if (missing(validation))
        validation = function (x) TRUE
    else
        stopifnot(inherits(validation, 'function') &&
                  length(formals(validation)) > 0)

    structure(as.list(environment()), class = 'sys$cmdline$opt')
}
```

```{r echo=FALSE}
`print.sys$cmdline$opt` = function (x, ...) {
    cat(sprintf("%s: [-%s|--%s] (default: %s) %s\n",
                x$variable_name,
                x$short,
                x$long,
                x$default,
                x$description))
    invisible(x)
}
```

## Test cases

```{r echo=FALSE, show='hide'}
# TODO: Implement
test = function (cmdline, spec, ...) {
    #parsed = parse_cmd(cmdline, spec)
    #testthat::expect_equals(parsed, list(...))
    spec
}
```

Display the help and exit.

```{r}
test('', NULL)
test('-h', NULL)
test('--help', NULL)
```

Collection of short arguments.

```{r}
test('-v',
     list(opt('v', 'verbose', FALSE, 'verbose logging')),
     list(verbose = TRUE))
test('-O2',
     list(opt('O', '', c(optimization = 1), 'optimization level')),
     list(optimization = 2))
# special case: we only allow “-v”, “-vv”, “-vvv” — needs to be validated separately.
test('-vvv', list(opt('v', '', c(verbose = ''), 'level of verbose logging',
                      function (value) value %in% c('', 'v', 'vv'))),
     list(verbose = 'vv'))
# The following doesn’t make sense but hey.
test('-vs', list(opt('v', 'verbose', FALSE, 'verbose logging'),
                 opt('s', 'silent', FALSE, 'silent output')),
     list(verbose = TRUE, silent = TRUE))
```

Equivalently, long options.

```{r}
test('--verbose',
     list(opt('v', 'verbose', FALSE, 'verbose logging')),
     list(verbose = TRUE))
test('--optim 2',
     list(opt('O', 'optim', c(optimization = 1), 'optimization level')),
     list(optimization = 2))
test('--optim=2',
     list(opt('O', 'optim', c(optimization = 1), 'optimization level')),
     list(optimization = 2))
test('--verbose --silent', list(opt('v', 'verbose', FALSE, 'verbose logging'),
                                opt('s', 'silent', FALSE, 'silent output')),
     list(verbose = TRUE, silent = TRUE))
```
