# Command line applications for R

```{r echo=FALSE}
system('stty columns 80')
script_name = 'calc.r'

strip = function (str)
    sub('^\\s*(.*?)\\s*$', '\\1', str)

inline = function (code)
    paste0('`', code, '`')

code = function (code)
    paste('```', code, '```', sep = '\n')

exec = function (cmd)
    strip(paste(suppressWarnings(system(cmd, intern = TRUE)), collapse = '\n'))

script = function (...) {
    cmd = paste(c(sprintf('./%s', script_name), unlist(list(...)), '2>&1'),
                collapse = ' ')
    exec(cmd)
}
```

This is a work in progress for a `sys` module that supports command line
applications in R. Inspired by Python, it implements a simple, opinionated
paradigm for implementing command line tools.

To ensure that the code of the command line application is itself reusable,
`sys$run` offers an entry point that is only invoked when the code is run
self-contained from the command line, not when it’s included as a module
(similar to Python’s `if __name__ == '__main__'`). The module also offers a
self-contained command line parser that supports largely the same syntax as
getopt and argparse, but with an R-like API.

This module supports writing tools using code such as the following:

```{r eval=FALSE, code=readLines(script_name)}
```

Saving the code as `r inline(script_name)`, making it executable and running it,
via

```bash
./calc.r -d 2 '10 / 3'
```

yields the output `r inline(script('-d 2 "10 / 3"'))`. Running it without
arguments or with invalid arguments displays an error message:

```{r echo=FALSE, results='asis'}
cat(code(script()))
```

Running it via `./calc.r --help` displays a formatted help message, listing the
available options, their description and default values:

```{r echo=FALSE, results='asis'}
cat(code(script('--help')))
```

## Status

This module already largely works but the API (in particular that of the command
line parser) are still subject to change. Use at your own risk.
